Puck Predictions: Unraveling the NHL Game Forecasting Riddle

### 2024-03-30 Latex/draft.tex
finished analysis and results for MCMC and accuracy

\begin{abstract}
    The goal of this project is simple: predict the outcomes of NHL games from any given state.
    As simple as the problem statement is, however, the solution is not so straightforward.
    To solve this problem, we will use a variety of machine learning techniques, including logistic regression,
    XGBoost, and ARIMA models. Additionally, we utilize a form of MCMC to simulate the outcomes of games from any given
    state. Our hypothesis is that we will be able to successfully predict the outcomes of NHL games with a high degree of accuracy
    using these tools.
\end{abstract}

@@

To simulate a game, we performed this Monte Carlo algorithm that acted like a random walk through the Markov Chain:
\begin{algorithm}[H]
    \caption{Simulation Algorithm}
    \begin{algorithmic}[1]
        \State $\text{time\_remaining} \gets 3600$
        \Comment{NHL games are 3600 seconds}
        \State $s_0 \gets "\#"$
        \Comment{The "\#" symbol represents the start of a game}
        \State $s_1 \gets "\#"$
        \State $s_2 \gets "\#"$
        %\State $\text{event\_counts} \gets \text{dictionary()}$

        \While{$\text{time\_remaining} > 0$}
            \State $\text{next\_state} \gets \text{sample from } \text{MC(curr\_state)}$
            %\State $\text{event\_counts}[\text{next\_state}] \mathrel{+}= 1$

@@

            \State $\text{event\_time} \gets \text{sample from } \text{KDE\_times()}$
            \State $\text{time\_remaining} \gets (\text{time\_remaining} - \text{event\_time})$
        \EndWhile
    \end{algorithmic}
    \label{alg:monte_carlo_algorithm}
\end{algorithm}
The KDE\_times() is a KDE model fit on the amount of seconds that transpired between 
each hockey event for all NHL games over the course of 13 years. This KDE model 
resembled a poisson distribution with $\lambda \approx 18$. During the course of 
the algorithm, we maintained a count of all of the events that transpired. Upon 
termination of the simulation, we compared home goals to away goals and termed 
a winner. If there was a tie, we would run the algorithm until another team scored0 
a goal, thus simulating overtime per NHL rules.

To predict a probability of a certain team winning a hockey game, we would run 
50 simulations with our algorithm, but we would initialize the starting states 
$(s_0, s_1, s_2)$ to be the most current events in the hockey game. Out of the 
50 simulations, we would compute 
$P(\text{home winning}) = \{\text{\# home simulation wins}\}/50$ and 
$P(\text{away winning}) = 1-P(\text{home winning})$.

To evaluate our simulation's effectiveness at accurately modeling actual hockey 
games, we combined a dataset with the final event counts for 1500 actual hockey 
games and for 1500 simulated games. We then performed PCA, t-SNE and UMAP 
dimensionality reductions using various perplexity and neighbor hyperparameters 
to see if these algorithms clustered the synthetic games and actual games in 
different clusters. As shown below in~\ref{fig:simulation_v_actual}, the simulated
games and actual games are all clustered together, thus demonstrating that our
simulation effectively emulates live NHL games.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{images/pca_tsne_umap_sim_act.png}
    \caption{PCA, t-SNE, and UMAP performed on a dataset with final event counts for 1500 simulated and 1289 actual NHL games. The joined grouping demonstrates that our simulation accurately emulates live NHL games}
    \label{fig:simulation_v_actual}
\end{figure}

@@

>>>>

\begin{abstract}
    The goal of this project is simple: predict the outcomes of NHL games from any given state during the game. To solve this problem we use three primary methods: Bayesian Network to train a DAG, XGBoost on game states, and an MCMC game simulator. We use each of these methods to generate win probabilities for each time step throughout various NHL games, thus emulating a live win probability. Finally, we analyze the accuracy of these three methods and considered ethical implications of our results.
\end{abstract}

@@

To simulate a game, we performed this Monte Carlo algorithm that acted like a random walk through the Markov Chain (Note that the KDE\_times() is a KDE model fit on the amount of seconds that transpired between 
each hockey event for all NHL games over the course of 13 years):
\begin{algorithm}[H]
    \caption{Simulation Algorithm}
    \begin{algorithmic}[1]
        \State $\text{time\_remaining} \gets 3600$
        \Comment{NHL games are 3600 seconds}
        \State $s_0 \gets "\#"$
        \Comment{The "\#" symbol represents the start of a game}
        \State $s_1 \gets "\#"$
        \State $s_2 \gets "\#"$
        \State $\text{event\_counts} \gets \text{dictionary()}$

        \While{$\text{time\_remaining} > 0$}
            \State $\text{next\_state} \gets \text{sample from } \text{MC(curr\_state)}$
            \State $\text{event\_counts}[\text{next\_state}] \mathrel{+}= 1$

@@

            \State $\text{event\_time} \gets \text{sample from } \text{KDE\_times()}$
            \State $\text{time\_remaining} \gets (\text{time\_remaining} - \text{event\_time})$
        \EndWhile
    \end{algorithmic}
    \label{alg:monte_carlo_algorithm}
\end{algorithm}

To predict a team's winning probability, we would simulate 50 hockey games with the initial starting states $(s_0, s_1, s_2)$ set to the most current events in the hockey game. By looking at each games
final event counts, we could compute each game's winner. Then we computed the 
probability as $P(\text{home winning}) = \{\text{\# home simulation wins}\}/50$ and 
$P(\text{away winning}) = 1-P(\text{home winning})$.

@@

\subsection{MCMC Simulator}

To evaluate our simulation's effectiveness at accurately modeling actual hockey 
games, we combined a dataset with the final event counts for 1500 actual hockey 
games and for 1500 simulated games. We then performed two experiments. First, We 
trained a KMeans, Logistic Regression, RandomForest and XGBoost model on this 
dataset with default parameters, to see 
if they could classify games as simulated or real. With a 70-30 train-test split, 
the accuracies on the test set were 47.9\%, 48.5\%, 67.5\% and 69.3\% respectively. 
Second, we fit and transformed this dataset using PCA, t-SNE and UMAP 
dimensionality reductions using various perplexity and neighbor hyperparameters to see 
if these algorithms clustered the synthetic games and actual games in 
different clusters. As shown below in ~\ref{fig:simulation_v_actual}, the simulated
games and actual games are all clustered together. From these two experiments, we conclude
that our simulator effectively models live NHL games.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{images/pca_tsne_umap_sim_act.png}
    \caption{PCA, t-SNE, and UMAP performed on a dataset with final event counts for 1500 simulated and 1500 actual NHL games. The joined grouping demonstrates that our simulation accurately emulates live NHL games. Note that we performed the t-SNE and UMAP on many different values of perplexity and neighbors, but all of the plots looked the same, so we just show this one}
    \label{fig:simulation_v_actual}
\end{figure}

\subsection{Method Comparison}
Each of the methods described above had one general purpose: accurately predict the outcome of a hockey game regardless of the point in the game. To evaluate our models, we randomly selected 50 games outside of the training data, and for various time steps throughout these games we had each model calculate a win probability. We then used the probability threshold of 0.5 to determine whether or not each model predicted if the home team won or lost. We then computed each model's accuracy across all 50 games for each of the time steps.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{images/game_accuracies.png}
    \caption{Each of our model's predictive accuracy over the course of 50 games}
    \label{fig:game_accuracies}
\end{figure}

In figure ~\ref{fig:game_accuracies} we note that each model increased in accuracy as the game progressed. We also notice that the MCMC simulation method appears to have the highest accuracy regardless of the point in the game while XGBoost appears to be the least accurate. We infer that the XGBoost models were overfit, since they appeared to perform very well on games within the training set. Also, during experimentation with the models, we noticed that the MCMC simulator and the Bayesian Network tended to produce higher probabilities for the away team winning, and the XGBoost model tended to give higher probabilities to the home team winning. The MCMC and Bayesian Network are structured to model the flow of events, so it appears that they picked up on a more predictive flow in how away teams play.
